---
title: "k-mer data analysis"
author: "Hongxuan Zhai and Julia Fukuyama"
date: "9/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

In this tutorial, we show how to use the defined functions to analyze the real data set and to generate the plot in the paper. The real data is a gut microbiome data set (Fukuyama et al., 2017) which focuses on the stability, resilience, and response to perturbation of the bacterial communities in the human gut. In this data set, there are 2611 taxa and 419 samples, and each taxon is represented by a sequence of length 233. This tutorial also includes the filtering and pre-processing steps to deal with rare taxa in the samples.



## Import Libraries

```{r required}
library(tidyverse)
library(phyloseq)
library(adaptiveGPCA)
library(DistatisR)
library(dplyr)
library(ggplot2)
library(vegan)
library(ggpubr)
library(kableExtra)
library(kernlab)
library(reshape2)
library(viridis)
library(patchwork)
library(treeDA)
library(treeio)
```

## Load Real Data
```{r real_data}
load("../data/phyloseq_objects.RData")
```


## Functions

```{r functions}
eksDist <- function(otutab, MMt, D = rep(1, nrow(otutab))) {
    Qeig = eigen(MMt)
    X = otutab
    X = scale(X, scale = FALSE)
    out.gpca = adaptiveGPCA:::gpcaEvecs(X,
                                        evecs = Qeig$vectors,
                                        evals = Qeig$values,
                                        D = D, k = min(nrow(X), ncol(X)))
    Uscaled = sweep(out.gpca$U, 2, STATS = out.gpca$lambda, FUN = "*")
    return(dist(Uscaled, method = "euclidean"))
}

mpqDist <- function(otutab, Q, r = 1, D = rep(1, nrow(otutab))) {
    Qeig = eigen(Q)
    X = otutab
    X = scale(X, scale = FALSE)
    if(r < 1 && r > 0) {
        out.gpca = adaptiveGPCA:::gpcaEvecs(X,
                                            evecs = Qeig$vectors,
                                            evals = (rep(r^(-1), ncol(X)) + (1-r)^(-1) * Qeig$values^(-1))^(-1),
                                            D = D, k = min(nrow(X), ncol(X)))
    } else if(r == 1) {
        out.gpca = adaptiveGPCA:::gpcaEvecs(X,
                                            evecs = Qeig$vectors,
                                            evals = Qeig$values,
                                            D = D, k = min(nrow(X), ncol(X)))
    } else if(r == 0){
        out.gpca = adaptiveGPCA:::gpcaEvecs(X,
                                            evecs = Qeig$vectors,
                                            evals = rep(1, ncol(X)),
                                            D = D, k = min(nrow(X), ncol(X)))
    }
    Uscaled = sweep(out.gpca$U, 2, STATS = out.gpca$lambda, FUN = "*")
    return(dist(Uscaled, method = "euclidean"))
}

#' Function to filtering out rare RSVs
#' 
#' @param phylo_obj a phyloseq object
#' @param count minimum number of occurance for a RSV used in filtering
#' @param portion used in filtering
#' @param psudo_count number to be added when 0 is prsent
#' 
#' @return a phyloseq object with pruned phylogeny 
filter_rare <- function(phylo_obj, count, portion, psudo_cnt = 0) {
  require(phyloseq)
  res = phyloseq::filter_taxa(phylo_obj, function(x) sum(x > count) > 
                                (portion*length(x)), TRUE)
  if (psudo_cnt) {
    res = phyloseq::transform_sample_counts(res, function(x) x+psudo_cnt)}
  return(res)
}

```

## Simulation Data Analysis

In this section, we use the simulated results to make plots for the eigenstructure of 
different tree. We first load three different tree types discussed in the paper (complete balanced binary tree, comb tree, real tree from the gut microbiome study (Fukuyama et al., 2017)).

```{r}
bt <- treeio::read.beast.newick("../data/btree.tree") # binary tree
ct <- treeio::read.beast.newick("../data/ctree.tree") # comb tree
rt <- treeio::read.beast.newick("../data/real_tree.xml") # real tree
```
Next, we read in the simulated eigenvalues and eigenvectors using different pholygenetic tree structure. The columns indicate eigenvalues/eigenvectors under different k in k-mer.
```{r}
b_eigvals <- read.csv("../data/b_eigvals_10000.csv", header = TRUE) # eigenvalues for binary tree
c_eigvals <- read.csv("../data/c_eigvals_10000.csv", header = TRUE) # eigenvalues for comb tree
r_eigvals <- read.csv("../data/r_eigvals_10000.csv", header = TRUE) # eigenvalues for real tree
```
Then we plot the relative weights for each eigenvalues, where the weights for eigenvalue $\lambda_i$ is defined as 
$$\frac{\lambda_i}{\sum_{m=0}^p\lambda_m}$$

```{r}
b_eigvals  = as.data.frame(lapply(b_eigvals, function(x) x/sum(x)))
c_eigvals  = as.data.frame(lapply(c_eigvals, function(x) x/sum(x))) 
r_eigvals  = as.data.frame(lapply(r_eigvals, function(x) x/sum(x))) 
colnames(b_eigvals) <- c('k=1', 'k=2', 'k=3', 'k=4', 'k=5', 'k=6', 'k=7')
colnames(c_eigvals) <- c('k=1', 'k=2', 'k=3', 'k=4', 'k=5', 'k=6', 'k=7')
colnames(r_eigvals) <- c('k=1', 'k=2', 'k=3', 'k=4', 'k=5', 'k=6', 'k=7')
b_eigvals = b_eigvals %>% 
  rownames_to_column("index") %>%
  gather(k, values, -index)
c_eigvals = c_eigvals %>% 
  rownames_to_column("index") %>%
  gather(k, values, -index)
r_eigvals = r_eigvals %>% 
  rownames_to_column("index") %>%
  gather(k, values, -index)
b_eigvals$tree = rep('binary', 224)
c_eigvals$tree = rep('comb', 224)
r_eigvals$tree = rep('real', 350)
b_eigvals$index <- as.numeric(b_eigvals$index)
c_eigvals$index <- as.numeric(c_eigvals$index)
r_eigvals$index <- as.numeric(r_eigvals$index)

bp1 <- ggplot(b_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + theme_bw() +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
  xlab("order") + ylab("value") + ggtitle("binary tree")

bp2 <- ggplot(b_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + coord_cartesian(xlim = c(2, 9), ylim = c(0, 40)) + theme_bw() +
  theme(legend.position = "none", panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + xlab("") + ylab("")

bp_zoom = ggplotGrob(bp2)
bp = bp1 + annotation_custom(grob = bp_zoom, xmin = 10, xmax = 32, ymin = 20, ymax = 360)

cp1 <- ggplot(c_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + theme_bw() +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))+
  xlab("order") + ylab("value") + ggtitle("comb tree")

cp2 <- ggplot(c_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + coord_cartesian(xlim = c(2, 9), ylim = c(0,30)) + theme_bw() +
  theme(legend.position = "none", panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + xlab("") + ylab("")

cp_zoom = ggplotGrob(cp2)
cp = cp1 + annotation_custom(grob = cp_zoom, xmin = 7, xmax = 32, ymin = 20, ymax = 350)

rp1 <- ggplot(r_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + theme_bw() +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))+
  xlab("order") + ylab("value") + ggtitle("real tree")

rp2 <- ggplot(r_eigvals, aes(index, values, group = k)) + geom_path(aes(color = k)) +
  geom_point(aes(color = k)) + coord_cartesian(xlim = c(2, 9), ylim = c(0, 30)) + theme_bw() +
  theme(legend.position = "none", panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + xlab("") + ylab("")

rp_zoom = ggplotGrob(rp2)
rp = rp1 + annotation_custom(grob = rp_zoom, xmin = 7, xmax = 48, ymin = 50, ymax = 700)

p_eig <- ggarrange(bp, cp, rp,
                   ncol = 3, nrow = 1, common.legend = TRUE, legend = 'bottom')
```

## Real Data Analysis

We filter out taxa that were not seen with more than 3 counts in at least 20% of our samples.
Compute the k-mer distances and MPQ distances on the log-transformed data:
```{r}
pruned_phylo = filter_rare(cc, 3, 0.2)
pruned_phylo_log = pruned_phylo
otu_table(pruned_phylo_log) = log(1 + otu_table(pruned_phylo))
pp = processPhyloseq(pruned_phylo_log)
k_vals = c(1:10, 20, 30, 40, 50, 75)
MMt_list = lapply(k_vals, function(k) {
    sk = stringdot(type = "spectrum", length = k, normalized = FALSE)
    kernel_matrix = kernelMatrix(sk, taxa_names(pruned_phylo))
})
r_vals = c(0, .01, .1, .5, .9, .95, .99, .999, 1)
mpq_distances = lapply(r_vals, function(r) mpqDist(pp$X, pp$Q, r = r))
eks_distances = lapply(MMt_list, function(MMt) eksDist(pp$X, MMt))
```



Perform distatis on the distances to see the relationships among them:
```{r}
distance_list = c(mpq_distances, eks_distances)
distance_types = rbind(
    data.frame(family = "MPQ", param = r_vals),
    data.frame(family = "EKS", param = k_vals)
)
distance_array = array(dim = c(nsamples(pruned_phylo), nsamples(pruned_phylo), length(distance_list)))
for(i in 1:length(distance_list)) {
    distance_array[,,i] = as.matrix(distance_list[[i]])
}
out_distatis = distatis(distance_array)
```

Look at the relationships between the MPQ distances and the k-mer distances:
```{r}
labs = sprintf("Axis %i: %4.2f%%", 1:length(distance_list),
    out_distatis$res4Cmat$eigValues / sum(out_distatis$res4Cmat$eigValues) * 100)
plotting = data.frame(distance_types, sweep(scale(out_distatis$res4Cmat$eigVector, center = TRUE), MARGIN = 2, STAT = sqrt(out_distatis$res4Cmat$eigValues), FUN = "*"))
pdf("mpq-eks-distatis.pdf", width = 6, height = 4.5)
ggplot() +
    geom_point(aes(x = dim.1, y = dim.2, color = param, shape = 16),
               data = subset(plotting, family == "EKS"), size = 3) + 
    geom_point(aes(x = dim.1, y = dim.2, fill = param, shape = 22),
               data = subset(plotting, family == "MPQ"), size = 3) +
    xlab(labs[1]) +  ylab(labs[2]) +
    scale_fill_gradientn("r", colours=viridis(n = 100, begin = .6, end = .35, option = "F")) +
    scale_color_gradientn("k",
                          colours=viridis(n = length(k_vals), begin = .35, end = .6, option = "G"),
                          breaks = k_vals[seq(1, length(k_vals), by=2)], trans = "log2") +
    scale_shape_identity("Family", breaks = c(22, 16), labels = c("MPQ", "EKS"), guide = "legend")
dev.off()
```


Look at the RV coefficient matrix:
```{r}
rv_mat = out_distatis$res4Cmat$C
pca_idx = which(distance_types$family == "MPQ" & distance_types$param == 0)
names = apply(distance_types, 1, function(fp) {
    family = fp[1]
    param = as.numeric(fp[2])
    if(family == "MPQ") {
        return(sprintf("%s %.2f", family, param))
    } else {
        return(sprintf("%s %2i", family, param))
    }
})
#names[pca_idx] = "Euclidean"
rownames(rv_mat) = colnames(rv_mat) = names
n_r = length(r_vals)
n_k = length(k_vals)
mpq_eks_rv = melt(rv_mat[1:n_r, (n_r+1):(n_r+n_k)], varnames = c("MPQ", "EKS"))
closest_matches = mpq_eks_rv %>% group_by(EKS) %>% summarise(max_mpq = MPQ[which.max(value)])
pdf("mpq-eks-rv.pdf", width = 7, height = 4)
ggplot(mpq_eks_rv) +
    geom_tile(aes(x = MPQ, y = EKS, fill = value)) +
    geom_tile(aes(x = max_mpq, y = EKS), fill = rgb(0,0,0,0), size = .5, color = "black", data = closest_matches) +
    scale_fill_viridis("RV coef") + theme(axis.text.x = element_text(angle = 90, vjust = .5)) + xlab("") + ylab("") + coord_flip()
dev.off()
```

Look at the embeddings in the first two dimensions for a subset of the distances:
```{r}
embedding_list = lapply(distance_list, cmdscale)
plots = lapply(embedding_list, function(x) {
    colnames(x) = paste0("Axis", 1:ncol(x))
    p = ggplot(data.frame(x, sample_data(cc))) +
        geom_point(aes(x = Axis1, y = Axis2, color = Subject), size = .8) +
        scale_shape("Time") +
        theme(
              axis.ticks.x = element_blank(), axis.text.x = element_blank(),
              axis.ticks.y = element_blank(), axis.text.y = element_blank(),
              axis.title.x = element_text(size = 12),
              axis.title.y = element_text(size = 12))
    return(p)
})
plots = mapply(FUN = function(p, name) p + ggtitle(name), plots, rownames(rv_mat), SIMPLIFY = FALSE)
to_plot = c(9, 6, 5, 4, 1, 10, 11, 12, 13, 14, 17, 19, 21, 23, 24)
pdf("mpq-eks-embedding.pdf", width = 8, height = 5)
wrap_plots(plots[to_plot], ncol = 5) + plot_layout(guides = "collect")
dev.off()
```

```{r}
for_table = c(1, 4, 5, 6, 9, 10, 11, 12, 13, 14, 17, 19, 21, 23, 24)
kbl(round(rv_mat[for_table, for_table], digits = 2), booktabs = TRUE, format = "latex")
```


## Repeat with a different transformation (clr)

```{r}
pruned_phylo = filter_rare(cc, 3, 0.2)
pruned_phylo_clr = pruned_phylo
log_trans_otu_table = log(1 + otu_table(pruned_phylo))
clr_trans_otu_table = sweep(log_trans_otu_table, MARGIN = 1, STATS = rowMeans(log_trans_otu_table), FUN = '-')
otu_table(pruned_phylo_clr) = otu_table(clr_trans_otu_table, taxa_are_rows = FALSE)
pp = processPhyloseq(pruned_phylo_clr)
k_vals = c(3:10, 20, 30, 40, 50, 75, 150)
MMt_list = lapply(k_vals, function(k) {
    sk = stringdot(type = "spectrum", length = k, normalized = FALSE)
    kernel_matrix = kernelMatrix(sk, taxa_names(pruned_phylo))
})
r_vals = c(0, .01, .1, .5, .9, .95, .99, .999, 1)
mpq_distances = lapply(r_vals, function(r) mpqDist(pp$X, pp$Q, r = r))
eks_distances = lapply(MMt_list, function(MMt) eksDist(pp$X, MMt))
```

```{r}
distance_list = c(mpq_distances, eks_distances)
distance_types = rbind(
    data.frame(family = "MPQ", param = r_vals),
    data.frame(family = "EKS", param = k_vals)
)
distance_array = array(dim = c(nsamples(pruned_phylo), nsamples(pruned_phylo), length(distance_list)))
for(i in 1:length(distance_list)) {
    distance_array[,,i] = as.matrix(distance_list[[i]])
}
out_distatis = distatis(distance_array)

labs = sprintf("Axis %i: %4.3f%%", 1:length(distance_list),
    out_distatis$res4Cmat$eigValues / sum(out_distatis$res4Cmat$eigValues) * 100)
plotting = data.frame(distance_types, sweep(scale(out_distatis$res4Cmat$eigVector, center = TRUE), MARGIN = 2, STAT = sqrt(out_distatis$res4Cmat$eigValues), FUN = "*"))
pdf("mpq-eks-distatis-clr.pdf", width = 6, height = 4.5)
ggplot() +
    geom_point(aes(x = dim.1, y = dim.2, color = param, shape = 16),
               data = subset(plotting, family == "EKS"), size = 3) + 
    geom_point(aes(x = dim.1, y = dim.2, fill = param, shape = 22),
               data = subset(plotting, family == "MPQ"), size = 3) +
    xlab(labs[1]) +  ylab(labs[2]) +
    scale_fill_gradientn("r", colours=viridis(n = 100, begin = .6, end = .35, option = "F")) +
    scale_color_gradientn("k",
                          colours=viridis(n = length(k_vals), begin = .35, end = .6, option = "G"),
                          breaks = k_vals[seq(1, length(k_vals), by=2)], trans = "log2") +
    scale_shape_identity("Family", breaks = c(22, 16), labels = c("MPQ", "EKS"), guide = "legend")
dev.off()

rv_mat = out_distatis$res4Cmat$C
pca_idx = which(distance_types$family == "MPQ" & distance_types$param == 0)
names = apply(distance_types, 1, function(fp) {
    family = fp[1]
    param = as.numeric(fp[2])
    if(family == "MPQ") {
        return(sprintf("%s %.3f", family, param))
    } else {
        return(sprintf("%s %2i", family, param))
    }
})
#names[pca_idx] = "Euclidean"
rownames(rv_mat) = colnames(rv_mat) = names
n_r = length(r_vals)
n_k = length(k_vals)
mpq_eks_rv = melt(rv_mat[1:n_r, (n_r+1):(n_r+n_k)], varnames = c("MPQ", "EKS"))
closest_matches = mpq_eks_rv %>% group_by(EKS) %>% summarise(max_mpq = MPQ[which.max(value)])
pdf("mpq-eks-rv-clr.pdf", width = 7, height = 4)
ggplot(mpq_eks_rv) +
    geom_tile(aes(x = MPQ, y = EKS, fill = value)) +
    geom_tile(aes(x = max_mpq, y = EKS), fill = rgb(0,0,0,0), size = .5, color = "black", data = closest_matches) +
    scale_fill_viridis("RV coef") + theme(axis.text.x = element_text(angle = 90, vjust = .5)) + xlab("") + ylab("") + coord_flip()
dev.off()

embedding_list = lapply(distance_list, cmdscale)
plots = lapply(embedding_list, function(x) {
    colnames(x) = paste0("Axis", 1:ncol(x))
    p = ggplot(data.frame(x, sample_data(cc))) +
        geom_point(aes(x = Axis1, y = Axis2, color = Subject), size = .8) +
        scale_shape("Time") +
        theme(
              axis.ticks.x = element_blank(), axis.text.x = element_blank(),
              axis.ticks.y = element_blank(), axis.text.y = element_blank(),
              axis.title.x = element_text(size = 12),
              axis.title.y = element_text(size = 12))
    return(p)
})


plots = mapply(FUN = function(p, name) p + ggtitle(name), plots, rownames(rv_mat), SIMPLIFY = FALSE)
to_plot = c(9, 6, 5, 4, 1,
            10, 11, 12, 17, 18, 19, 20, 21, 22, 23)
pdf("mpq-eks-embedding-clr.pdf", width = 8, height = 5)
wrap_plots(plots[to_plot], ncol = 5) + plot_layout(guides = "collect")
dev.off()

for_table = c(1,4,5,6,9,
            10, 11, 12, 17, 18, 19, 20, 21, 22, 23)
kbl(round(rv_mat[for_table,for_table], digits = 2), booktabs = TRUE, format = "latex")
```
